How General Health Index (GHI) performs better to predict health levels

### 2024-12-03 paper/paper.qmd
everything so far

library(tidyverse)
library(janitor)
library(arrow)
library(dplyr)
```

@@

```{r}
#| warning: false
#| message: false
#| echo: false
body_mass_data_raw <- read_parquet("../data/raw_data/body_mass_data.parquet")
body_mass_clean <- body_mass_data_raw %>%
  clean_names() %>%
  select(gender_m1f2, age, height_cm, w_kg, bmi, fat_mass_kg, fm, ffm_kg, bone_mass_kg, muscle_mass_kg) %>%
  mutate(
    bmi_category = case_when(
      bmi < 18.5 ~ "Underweight",
      bmi >= 18.5 & bmi <= 24.9 ~ "Normal weight",
      bmi >= 25.0 & bmi <= 29.9 ~ "Overweight",
      bmi >= 30.0 & bmi <= 34.9 ~ "Obese - Class I",
      bmi >= 35.0 & bmi <= 39.9 ~ "Obese - Class II",
      bmi >= 40.0 ~ "Obese - Class III"
    ))

# Data Section: Part 1: What is the difference between an underweight / athletic person / normal person:
bmi_under_25 <- body_mass_clean %>%
  filter(bmi < 25)

ggplot(bmi_under_25, aes(x = bmi, y = fm)) +
  geom_point(aes(color = bmi_category), alpha = 0.7) +
  labs(
    title = "BMI vs. Fat Mass by BMI Category",
    x = "BMI",
    y = "Fat Mass (kg)",
    color = "BMI Category"
  ) +
  theme_minimal()

```

### The difference between a normal and overweight person:
```{r}
#| warning: false
#| message: false
#| echo: false

bmi_normal_over <- body_mass_clean %>%
  filter(18.5 < bmi & bmi < 30)

# this is what i chose
ggplot(bmi_normal_over, aes(x = bmi, y = fm)) +
  geom_point(aes(color = bmi_category), alpha = 0.7) +
  labs(
    title = "BMI vs. Fat Mass by BMI Category",
    x = "BMI",
    y = "Fat Mass (kg)",
    color = "BMI Category"
  ) +
  theme_minimal()
```

@@

# this is what i chose
ggplot(body_mass_clean, aes(x = bmi, y = fm)) +
  geom_point(aes(color = bmi_category), alpha = 0.7) +
  geom_vline(xintercept = c(18.5, 25, 30), linetype = "dashed", color = "black", linewidth = 0.5, alpha = 0.5) +
  labs(
    title = "BMI vs. Fat Mass by BMI Category",
    x = "BMI",
    y = "Fat Mass (kg)",
    color = "BMI Category"
  ) +
  theme_minimal()
```

@@

@@

@@

@@

@@

Where:
- \( y \) is the **fat percentage category** (e.g., categories like 'low', 'medium', 'high').
- \( P(y = j) \) is the probability that the fat percentage falls into category \( j \).
- \( P(y = \text{ref}) \) is the reference category.
- \( \beta_{0j}, \beta_{1j}, \dots \) are the coefficients to be estimated for each category \( j \).

@@

# Results

Our results are summarized in @tbl-modelresults. Our model's results and interpretation:

library(caret)

# Ensure actual categories are factors
# actual_categories <- factor(cleaned_data_model$fat_percentage_category, 
#                             levels = c("high", "low", "moderate", "risk"))
# 
# # Ensure predicted categories are factors
# predicted_categories <- factor(predict(measurements_model, newdata = cleaned_data_model, type = "class"),
#                                 levels = c("high", "low", "moderate", "risk"))
# 
# # Now create the confusion matrix
# 
# 
# 
# conf_matrix <- confusionMatrix(predicted_categories, actual_categories)
# print(conf_matrix)
# accuracy <- mean(predicted_categories == actual_categories)
# print(accuracy)
# 
# library(pROC)
# 
# probabilities <- predict(measurements_model, newdata = cleaned_data_model, type = "probs")
# 
# # Convert true categories to a binary format for ROC computation (one-vs-all approach)
# roc_result <- multiclass.roc(actual_categories, probabilities)
# 
# # Plot ROC curve
# plot(roc_result)

```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

```

@@

## Third discussion point

## Weaknesses and next steps
- the BodyM dataset does not include age. An older person has lower metabolism [] and a higher chance of collecting fat on their abdomen.

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
```

>>>>

library(tidyverse)
library(janitor)
library(arrow)
library(dplyr)
library(caret)
library(knitr)

# reading the final datasets:
cleaned_data_model <- read_parquet(file = here::here("data/analysis_data/measurements_analysis_data.parquet"))
body_mass_model <- read_parquet(file = here::here("data/analysis_data/body_mass_data.parquet"))
```

@@

```{r}
#| warning: false
#| message: false
#| echo: false
body_mass_data_raw <- read_parquet("../data/raw_data/body_mass_data.parquet")
body_mass_clean <- body_mass_data_raw %>%
  clean_names() %>%
  select(gender_m1f2, age, height_cm, w_kg, bmi, fat_mass_kg, fm, ffm_kg, bone_mass_kg, muscle_mass_kg) %>%
  mutate(
    bmi_category = case_when(
      bmi < 18.5 ~ "Underweight",
      bmi >= 18.5 & bmi <= 24.9 ~ "Normal weight",
      bmi >= 25.0 & bmi <= 29.9 ~ "Overweight",
      bmi >= 30.0 & bmi <= 34.9 ~ "Obese - Class I",
      bmi >= 35.0 & bmi <= 39.9 ~ "Obese - Class II",
      bmi >= 40.0 ~ "Obese - Class III"
    ))
```
# Data Section: Part 1: What is the difference between an underweight / athletic person /

@@

# this is what i chose
ggplot(body_mass_model, aes(x = bmi, y = fm)) +
  geom_point(aes(color = bmi_category), alpha = 0.7) +
  geom_vline(xintercept = c(18.5, 25, 30), linetype = "dashed", color = "black", linewidth = 0.5, alpha = 0.5) +
  labs(
    title = "BMI vs. Fat Mass by BMI Category",
    x = "BMI",
    y = "Fat Mass (kg)",
    color = "BMI Category"
  ) +
  theme_minimal()
```

@@

```{r}
# measurements_tibble <- as_tibble(cleaned_data_model)
# kable(measurements_tibble, caption = "Example Tibble Output")
```

@@

cleaned_data_model$bmi_category <- factor(cleaned_data_model$bmi_category, 
                            levels = c("Underweight", "Normal", "Overweight", "Obese"))
ggplot(cleaned_data_model, aes(x = gender, fill = bmi_category)) +
  geom_bar(position = "fill") +  # 'fill' makes it a proportion plot
  ylab("Proportion") +
  xlab("Gender") +
  labs(title = "Proportions of BMI Categories by Gender") +
  scale_y_continuous(labels = scales::percent) +  # Converts y-axis to percentage format
  theme_minimal()
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
cleaned_data_model$gender <- factor(cleaned_data_model$gender, labels = c("Female", "Male"))
ggplot(cleaned_data_model, aes(x = gender, y = waist_hip_ratio, fill = gender)) +
  geom_boxplot() +
  ylab("Waist-to-Hip Ratio") +
  xlab("Gender") +
  labs(title = "Distribution of Waist-to-Hip Ratio by Gender") +
  theme_minimal() +
  facet_wrap(~fat_percentage_category) + 
  scale_fill_manual(values = c("female" = "lightblue", "male" = "lightgreen"))
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
cleaned_data_model$gender <- factor(cleaned_data_model$gender, labels = c("Female", "Male"))

cleaned_data_model$bmi_category <- factor(cleaned_data_model$bmi_category, 
                            levels = c("Underweight", "Normal", "Overweight", "Obese"))
ggplot(cleaned_data_model, aes(x = gender, y = height_hip_ratio, fill = gender)) +
  geom_boxplot() +
  ylab("Waist-to-Hip Ratio") +
  xlab("Gender") +
  labs(title = "Distribution of Waist-to-Hip Ratio by Gender") +
  theme_minimal() +
  facet_wrap(~bmi_category) + 
  scale_fill_manual(values = c("female" = "lightblue", "male" = "lightgreen"))
```
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
# filtered_data <- subset(cleaned_data_model, bmi_category %in% c("Overweight", "Obese"))

# Create a combined boxplot to visualize differences in ankle and wrist for bmi_category = Overweight and Obese

cleaned_data_model$bmi_category <- factor(cleaned_data_model$bmi_category, 
                            levels = c("Underweight", "Normal", "Overweight", "Obese"))

long_data <- cleaned_data_model %>%
  gather(key = "Measurement", value = "Value", ankle, wrist)

# Create side-by-side boxplots for ankle and wrist measurements
ggplot(long_data, aes(x = bmi_category, y = Value, fill = bmi_category)) +
  geom_boxplot() +
  facet_wrap(~ Measurement) +  # This creates separate panels for 'ankle' and 'wrist'
  ylab("Measurement Value") +
  xlab("BMI Category") +
  labs(title = "Comparison of Ankle and Wrist Measurements for Overweight and Obese BMI Categories") +
  theme_minimal() +
  scale_fill_manual(values = c("Overweight" = "lightcoral", "Obese" = "lightsalmon")) +
  theme(legend.position = "none")
```

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
# Recode gender levels
body_mass_model %>%
  mutate(gender = factor(gender, levels = c(1, 2), labels = c("Male", "Female"))) %>%
  mutate(fat_percentage_category = factor(fat_percentage_category,
                                          levels = c("Low", "Moderate", "High", "Extreme"))) %>%
  mutate(
    gender = as.factor(gender),
    fat_percentage_category = as.factor(fat_percentage_category)
  )

ggplot(body_mass_model, aes(x = gender, fill = fat_percentage_category)) +
  geom_bar(position = "fill") +  # 'fill' makes it a proportion plot
  ylab("Proportion") +
  xlab("Gender") +
  labs(title = "Proportions of BMI Categories by Gender") +
  scale_y_continuous(labels = scales::percent) +  # Converts y-axis to percentage format
  theme_minimal()
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
ggplot(body_mass_model, aes(x = weight, y = height, color = fm)) +
  geom_point(size = 3) +
  scale_color_viridis_c() + # Use a continuous color scale
  labs(title = "Height vs Weight with Fat Mass",
       x = "Weight",
       y = "Height",
       color = "Fat Mass (fm)") +
  theme_minimal()
```
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
ggplot(body_mass_model, aes(x = weight, y = height, color = fm, size = fm)) +
  geom_point(alpha = 0.7) +
  scale_color_viridis_c() +
  scale_size_continuous(range = c(1, 10)) +
  labs(title = "Height vs Weight with Fat Mass",
       x = "Weight",
       y = "Height",
       color = "Fat Mass (fm)",
       size = "Fat Mass (fm)") +
  theme_minimal()
```
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
ggplot(body_mass_model, aes(x = bmi, y = fm, color = gender)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ gender) +
  labs(title = "Fat Mass vs BMI by Gender",
       x = "BMI",
       y = "Fat Mass (fm)",
       color = "Gender") +
  theme_minimal()
```
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
body_mass_long <- body_mass_model %>%
  pivot_longer(cols = c(fm, bmi), names_to = "measure", values_to = "value")

# Dual y-axis plot
ggplot(body_mass_long, aes(x = weight, y = value, color = measure)) +
  geom_point(alpha = 0.5) + # Optional: Add points for raw data
  geom_smooth(se = FALSE, method = "loess", span = 0.5) + # Smoothed trendline
  labs(title = "Smoothed Trends of FM and BMI by Age",
       x = "weight",
       y = "Value",
       color = "Measure") +
  theme_minimal() +
  facet_wrap(~gender)

```
- talk about the crowding for men

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
# Combine both datasets
cleaned_test_extract <- cleaned_data_model %>%
  select(gender, bmi_category, fat_percentage_category)
body_mass_extracted <- body_mass_model %>%
  select(gender, bmi_category, fat_percentage_category)
cleaned_test_extract$dataset <- "Measurements"
body_mass_extracted$dataset <- "Body Data"
combined_data <- rbind(cleaned_test_extract, body_mass_extracted)

# Count occurrences of each fat_percentage_category per dataset
library(dplyr)
category_counts <- combined_data %>%
  group_by(dataset, fat_percentage_category) %>%
  summarise(Count = n(), .groups = "drop")

# Create a bar plot to visualize the differences
library(ggplot2)
ggplot(category_counts, aes(x = fat_percentage_category, y = Count, fill = dataset)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of Fat Percentage Categories Across Data1 and Data2",
       x = "Fat Percentage Category", y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("Measurements" = "lightblue", "Body Data" = "lightgreen"))

proportions <- combined_data %>%
  group_by(dataset, fat_percentage_category) %>%
  summarise(Count = n(), .groups = "drop") %>%
  mutate(Proportion = Count / sum(Count))

# Create a bar plot to visualize the proportions
ggplot(proportions, aes(x = fat_percentage_category, y = Proportion, fill = dataset)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Proportional Comparison of Fat Percentage Categories Across Data1 and Data2",
       x = "Fat Percentage Category", y = "Proportion") +
  scale_y_continuous(labels = scales::percent) +  # Show proportions as percentages
  theme_minimal() +
  scale_fill_manual(values = c("Measurements" = "lightblue", "Body Data" = "lightgreen"))
```

@@

<!-- Where: -->
<!-- \begin{align} -->
<!-- - \( y \) is the **fat percentage category** (e.g., categories like 'low', 'medium', 'high'). -->
<!-- - \( P(y = j) \) is the probability that the fat percentage falls into category \( j \). -->
<!-- - \( P(y = \text{ref}) \) is the reference category. -->
<!-- - \( \beta_{0j}, \beta_{1j}, \dots \) are the coefficients to be estimated for each category \( j \). -->
<!-- \end{align} -->

Where:
- \( y \) is the **fat percentage category**. The categories include low, moderate, high, and extreme. These equate low, moderate, high and extreme levels of body fat.
- \( P(y = j) \) is the probability that the fat percentage falls into category \( j \).
- \( P(y = \text{ref}) \) is the reference category.
- \( \beta_{0j}, \beta_{1j}, \dots \) are the coefficients to be estimated for each category \( j \).

@@

# Results
## Model Results for Measurements Model

Our results are summarized in @tbl-measurementmodelresults. Our model's results and interpretation:

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-measurementmodelresults
#| tbl-cap: "TODO"

# # Code chunk to import models
measurements_model_paper <- readRDS(file = here::here("models/measurements_model.rds"))
measurements_model_summary <- summary(measurements_model_paper)
measurements_summary_df <- as.data.frame(measurements_model_summary$coefficients)

# Create and style the table
kable(measurements_summary_df, caption = "Summary of Measurements Model")
```

## Model Results for Body Mass Model

Our results are summarized in @tbl-bodymass-modelresults. Our model's results and interpretation:

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-bodymass-modelresults
#| tbl-cap: "TODO"

# # Code chunk to import models
body_mass_model_paper <- readRDS(file = here::here("models/body_mass_model_bmi.rds"))
body_mass_model_summary <- summary(body_mass_model_paper)
body_mass_summary_df <- as.data.frame(body_mass_model_summary$coefficients)

# Create and style the table
kable(body_mass_summary_df, caption = "Summary of Measurements Model")
```

@@

## Second discussion point {#sec-second-point}

## Third discussion point {#sec-third-point}

## Weaknesses and next steps {#sec-weaknesses-point}
- the BodyM dataset does not include age. An older person has lower metabolism [] and a higher chance of collecting fat on their abdomen.

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false

# predictions <- as.factor(predict(measurements_model_paper, newdata = cleaned_data_model))
# reference <- as.factor(cleaned_data_model$fat_percentage_category)
# measurements_cm <- confusionMatrix(predictions, cleaned_data_model$fat_percentage_category)
# # Extract the confusion matrix
# measurements_conf_matrix <- as.data.frame(measurements_cm$table)
# measurements_conf_matrix_df <- as.data.frame(measurements_conf_matrix)
# rownames(measurements_conf_matrix_df) <- paste("Observed:", rownames(measurements_conf_matrix))
# colnames(measurements_conf_matrix_df) <- paste("Predicted:", colnames(measurements_conf_matrix))
# kable(measurements_conf_matrix_df, caption = "Confusion Matrix")
# 
# # Extract overall statistics
# overall_stats <- as.data.frame(t(measurements_cm$overall))
# kable(overall_stats, caption = "Overall Statistics")
# 
# residuals <- residuals(measurements_model_paper, type = "deviance")
# plot(residuals)
```
